\documentclass[xcolor=dvipsnames,10pt]{beamer}
\input{head.tex}

\title{\vskip 0.5in Day 2}
\subtitle{More on Portfolio Choice}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}[plain]
  \titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Review}
\subsection{}

\bfr\frametitle{Review of Portfolio Choice Problem}
\bi
\im Choose $\theta_1, \ldots, \theta_n$ to 
$$\max \; \mye\lb u\lp \sum_{i=1}^n \theta_i\tx_i \rp \rb \quad \text{subject to} \quad \sum_{i=1}^n p_i \theta_i = w\,.$$
\im Choose $\phi_1, \ldots, \phi_n$ to 
$$\max \;\mye\lb u\lp \sum_{i=1}^n \phi_i\tr_i \rp \rb \quad \text{subject to} \quad \sum_{i=1}^n \phi_i = w\,.$$
\im Choose $\pi_1, \ldots, \pi_n$ to 
$$\max \;\mye\lb u\lp w\sum_{i=1}^n \pi_i\tr_i \rp \rb \quad \text{subject to} \quad \sum_{i=1}^n \pi_i = 1\,.$$
Can define $\hat{u}(a) = u(wa)$ so objective is max $\mye[\hat{u}(\sum \pi_i \tr_i)]$.
\ei
\end{frame}



\bfr\frametitle{First-Order Condition}
\bi
\im If $p_i>0$ and $p_j>0$,
$$\mye\lb u'\lp \sum_{i=1}^n \theta_i\tx_i \rp (\tr_i -\tr_j)\rb = 0 $$
\im In words: marginal utility at the optimal wealth is orthogonal to excess returns.
\bi
\im A return is the payoff of a unit-cost portfolio.  
\im An excess return is the payoff of a zero-cost portfolio (for example, a difference of returns).
\ei 
\im Why?  Investing a little less in asset $j$ and a little more in asset~$i$ (or the reverse) cannot increase expected utility at the optimum.
\ei
\end{frame}




\section{Single Asset}
\subsection{}

\bfr\frametitle{A Single Risky Asset: \\Go Long if Risk Premium is Positive}
\bi
\im Assume a risk-free asset with return $R_f$ and a single risky asset with return $\tr$.  Set $\mu=\mye[\tr]$ and $\sigma^2 = \var(\tr)$.
\im Let $\phi=\,$ amount invested in risky asset, so $w_0-\phi$ is invested in risk-free asset.  
\im Date--1 wealth is 
$$\tw = (w_0-\phi)R_f + \phi \tr = w_0 R_f + \phi (\tr - R_f)$$
\im Claim: $\mu > R_f \Rightarrow\phi^*>0$ (by symmetry, $\mu<R_f \Rightarrow \phi^*<0$).
\ei
\end{frame}

\bfr\frametitle{Proof that $\phi>0$ when $\mu>R_f$}

\bi
\im 
We want to compare the utility of an investment $\phi>0$ to the utility of $\phi=0$.  That is, we want to compare $\mye[u(w_0R_f + \phi(\tr-R_f)]$ to $u(w_0R_f)$.  
\im 
Given $\phi$, expected wealth is
$\ow = w_0R_f + \phi(\mu-R_f)$.  Define $\tve = \tw - \ow$.
\im The investor will pay $\pi$ to avoid the gamble $\tve$ where 
$$u(\ow - \pi) = \mye[u(\ow + \tve)]\,.$$
\im The variance of $\tve$ is $\phi^2\sigma^2$, so by second-order risk aversion,
$$\pi \approx \frac{1}{2}\alpha(\ow)\phi^2\sigma^2 < (\mu-R_f)\phi$$
when $\phi>0$ and small, so
$$\mye[u(\tw)] = u(\ow-\pi) > u(\ow - (\mu-R_f)\phi) = u(w_0R_f)$$
\ei
\end{frame}


\bfr\frametitle{DARA Implies Risky Asset is a Normal Good}
\bi
\im Normal good: demand rises when income (wealth) rises.  Inferior: demand falls when income (wealth) rises.
\im A single risky asset with $\mu>R_f$ is a normal good if the investor has decreasing absolute risk aversion.
\im Proof:  The FOC is
$$\mye[u'(\tw)(\tr-R_f)] = 0$$ 
Differentiate it:
\begin{align*}
0 &= \frac{\D }{\D w_0} \mye[u'(w_0R_f + \phi(\tr-R_f))(\tr-R_f)]\\
&= \mye[u''(\tw)\{R_f + \phi'(w_0)(\tr-R_f)\}(\tr-R_f)]
\end{align*}
Rearrange as
$$\phi'(w_0) = - \frac{R_f \mye[u''(\tw)(\tr-R_f)]}{\mye[u''(\tw)(\tr-R_f)^2]}\,.$$
Can show: DARA $\,\Rightarrow\,$ $\phi' >0$.
\ei
\end{frame}

\section{Multiple Assets}
\subsection{}

\bfr\frametitle{Multiple Risky Assets}
\bi
\im $\tR=\,$ $n$--vector of risky asset returns
\im $\mu=\,$ $n$--vector of expected returns
\im $\phi=\,$ $n$--vector of investments in consumption good units
\im $\pi = (1/w_0)\phi$
\im $\iota=\,$ $n$--vector of 1's
\im $\Sigma = \,$ $n \times n$ covariance matrix, $\Sigma_{ij} = \cov(\tr_i,\tr_j)$
$$\Sigma = \mye[(\tr-\mu)(\tr-\mu)']$$
\im date--1 wealth $\tw = w_0R_f + \phi'(\tR - R_f\iota)$
\im expected wealth $\ow = w_0R_f + \phi'(\mu - R_f\iota)$
\im variance of wealth $\, = \phi'\Sigma\phi$.  Proof:
$$\mye[(\tw - \ow)^2] = \mye[\{\phi'(\tR - \mu)\}^2] = \mye[\phi'(\tr-\mu)(\tr-\mu)'\phi] = \phi'\Sigma\phi$$
\ei
\end{frame}

\bfr\frametitle{Eliminate Redundant Assets}
\bi
\im Each portfolio variance $\phi'\Sigma\phi \geq 0$ so a covariance matrix is positive semidefinite.
\im A positive semidefinite matrix is nonsingular if and only if it is positive definite.
\im Positive definiteness means all portfolio variances are positive, equivalently, there is no risk-free portfolio of risky assets.
\im If there is a risk-free portfolio of risky assets, one risky asset is a linear combination of the others and can be eliminated. Deleting it does not change investment opportunities.
\im By eliminating redundant assets, we can assume $\Sigma$ is nonsingular and positive definite.
\ei
\end{frame}

\bfr\frametitle{Diversification}
\bi
\im Portfolio variance is
$$\pi'\Sigma\pi = \sum_{i=1}^n \pi_i^2 \var(\tr_i) + 2 \sum_{i=1}^n \sum_{j=i+1}^n \pi_i\pi_j \cov(\tr_i,\tr_j)$$
\im We can generally make 
$\sum_{i=1}^n \pi_i^2 \var(\tr_i)$
small by diversifying, if there are many assets.
\im Suppose for example that the risky assets are uncorrelated and have the same variance $\sigma^2$ ($\Sigma = \sigma^2I$).  Then
$$\pi'\Sigma\pi = \sigma^2 \sum_{i=1}^n \pi_i^2$$
Among portfolios fully invested in risky assets ($\pi_i$ sum to 1), this variance is minimized at $\pi_i=1/n$ and
$$\pi'\Sigma\pi = \sigma^2 \sum_{i=1}^n \lp \frac{1}{n}\rp ^2 = \frac{\sigma^2}{n} \rightarrow 0 \quad \text{as $n \rightarrow \infty$}$$
\ei
\end{frame}

\section{CARA-Normal}
\subsection{}

\bfr\frametitle{CARA-Normal with Single Risky Asset}
Assume CARA utility $\mye[-\E^{-\alpha \tw}]$.  Assume $\tr\sim\,$ normal $(\mu,\sigma)$.  Then $\tw$ is normally distributed.  

Recall: If $\tx$ is normally distributed with mean $\mu_x$ and std dev $\sigma_x$, then
$$\mye[\E^{\tx}] = \E^{\mu_x + \sigma_x^2/2}$$

Given an investment $\phi$ in the risky asset, $-\alpha\tw$ is normal with mean $-\alpha w_0R_f - \alpha \phi(\mu-R_f)$ and std dev $\alpha\phi\sigma$.  Hence,
$$\mye[-\E^{-\alpha \tw}] = -\E^{-\alpha [w_0R_f + \phi(\mu-R_f) - \alpha\phi^2\sigma^2/2]}$$
Thus,
$$w_0R_f + \phi(\mu-R_f) - \alpha\phi^2\sigma^2/2$$
is the certainty equivalent (mean minus one-half risk aversion times variance).
\end{frame}

\bfr\frametitle{CARA-Normal cont.}
Optimal portfolio maximizes the certainty equivalent.  Therefore, the optimum is
$$\phi^* = \frac{\mu-R_f}{\alpha \sigma^2}$$

The optimal fraction of wealth to invest is
$$\pi^* = \frac{\mu-R_f}{(\alpha w_0)\sigma^2}$$

Usually assume $\alpha w_0$ (relative risk aversion at $w_0$) is between 2 and 10.
\end{frame}


\bfr\frametitle{CARA-Normal with Multiple Risky Assets}
\bi
\im Certainty equivalent is mean minus one-half risk aversion times variance:
$$w_0R_f + \phi'(\mu-R_f\iota) - \frac{1}{2}\alpha \phi'\Sigma\phi$$
\im FOC is
$$\mu-R_f\iota - \alpha\Sigma\phi = 0\,.$$
\im Optimum is
$$\phi^* = \frac{1}{\alpha}\Sigma^{-1}(\mu-R_f\iota)$$
Note no wealth effects.
\im Similar form to single risky asset case.  Optimum investment in each asset depends on its covariances with other assets unless $\Sigma$ is diagonal.
\ei
\end{frame}

\section{LRT Utility}
\subsection{}

\bfr\frametitle{LRT Utility and Wealth Expansion Paths}

\bi
\im CARA utility (even without normal returns) implies investments $\phi_i$ in risky assets are independent of initial wealth (absence of wealth effects).
\im CRRA utility implies portfolio weights $\pi_i$ are independent of initial wealth.
\im More generally, linear risk tolerance ($\tau(w) = A + Bw$) implies that the investment in each risky asset $i$ is
$$\phi_i = \xi_iA + \xi_iBR_fw_0$$
where $\xi_i$ does not depend on $A$ or $w_0$.
\bi
\im  CARA is the case $B=0$, so investments are independent of $w_0$ and inversely proportional to absolute risk aversion.
\im CRRA is the case $A=0$, so
$$\pi_i = \frac{\phi_i}{w_0} = \xi_iBR_f\,.$$
Remember $B=1/\rho$, so weights are inversely proportional to relative risk aversion.
\ei 
\ei

\end{frame}

\bfr\frametitle{Proof for Log Utility}
\bi
\im End-of-period wealth is (where $n=$ number of risky assets): 
$$\tw = w_0R_f + \sum_{i=1}^n \phi_i (\tilde{R}_i - R_f) = w_0 \left[R_f + \sum_{i=1}^n \pi_i (\tilde{R}_i - R_f)\right]\,.$$
\im Therefore, expected utility is 
$$\mye[\log(\tw)] \ = \ \log w_0 + \mye \left[\log \left(R_f + \sum_{i=1}^n \pi_i (\tilde{R}_i - R_f)\right)\right]\,.$$
\im The optimal $\pi_i$'s maximize
$$\mye \left[\log \left(R_f + \sum_{i=1}^n \pi_i (\tilde{R}_i - R_f)\right)\right]\,,$$
which does not depend on $w_0$.
\ei

\end{frame}

\end{document}

\bfr\frametitle{Exercises}
Graded: 2.4 

Recommended: 2.2, 2.3

In Exercise 2.3(a), ignore the hint ``Use the law of iterated expectations as in Section 1.5.''

\end{frame}

\bfr\frametitle{Exercise 2.6}
\bi
\im 
Assume time-additive CRRA utility:
$$v(c_0,c_1) = \frac{1}{1-\rho}c_0^{1-\rho} + \frac{\delta}{1-\rho}c_1^{1-\rho}$$
\im Standard definitions:
$$\text{Marginal Rate of Substitution} = \frac{\partial v/\partial c_0}{\partial v/\partial c_1}$$
$$\text{Elasticity of Intertemporal Substitution} = \frac{\D \log (c_1/c_0)}{\D \log MRS}$$
\im With time-additive CRRA utility, the EIS is the reciprocal of relative risk aversion.
\im Proof: Set $x=c_1/c_0$.  
Then MRS $\,= -\log \delta + \rho \log x$.  So, $\D \log \text{MRS} /\D \log x = \rho$.
\ei
\end{frame}

\section{Other}
\subsection{}



\blank
\blank
\blank
\blank
\blank

\end{document}

\bfr\frametitle{Two-Fund Separation with LRT Utility}
\bi
\im Suppose all investors have LRT utility with the same cautiousness parameter $B$.  
\im Investor $h$ invests
$$\phi^*_{hi} = \xi_i [A_h +  B R_f w_{h0}]$$
in risky asset~$i$.
The total investment of investor $h$ in risky assets is
$$\sum_{i=1}^n \phi^*_{hi} = \lp \sum_{i=1}^n \xi_i \rp [A_h + B R_f w_{h0}]$$
\im Of the total investment in risky assets, the fraction in asset~$i$ is
$$\frac{\xi_i}{\sum_{j=1}^n \xi_j}$$
\im This is the same for all investors.  Hence it is the market portfolio of risky assets.  All investors hold a combination of the risk-free asset and the market portfolio.
\ei
\end{frame}

\section{Time-Additive Utility}
\subsection{}

\bfr\frametitle{Time-Additive Utility and the Euler Equation}
\bi
\im Date--0 and date--1 consumption.  Utility function $v(c_0,c_1)$.  Assume time-additive utility
$$v(c_0,c_1) = u(c_0) + \delta u(c_1)$$
\im Consumption/investment problem:  choose $c_0, \phi_1, \ldots, \phi_n$ to 
$$\max \;u(c_0) +  \delta \mye\lb  u\lp \sum_{i=1}^n \phi_i\tr_i \rp \rb \quad \text{subject to} \quad c_0 + \sum_{i=1}^n \phi_i = w\,.$$
\im FOC: $u'(c_0) = \lambda$ and 
$$(\forall \,i) \quad  \mye\lb \delta u'\lp \sum_{i=1}^n \phi_i\tr_i \rp \tr_i\rb = \lambda$$
\im So
$$(\forall \,i) \quad \mye\lb \frac{\delta u'\lp \sum_{i=1}^n \phi_i\tr_i \rp}{u'(c_0)} \tr_i\rb = 1$$
\ei
\end{frame}

\section{LRT}
\subsection{}

\bfr\frametitle{Linear Risk Tolerance}
\bi
\im Assume no labor income.  Assume LRT $\tau(w) = A + B w$.  Assume a risk-free asset exists.
\im Then optimal investment in each risky asset (in consumption good units) is
$$\phi^*_i = \xi_i [A +  B R_f w_0]$$
where $\xi_i$ does not depend on $A$ or on $w_0$.  
\im Investors with LRT and the same cautiousness parameter have parallel wealth expansion paths.
\im CARA: $\phi^*_i = \xi_i A$ (no wealth effects)
\im CRRA: $\pi^*_i = \xi_i B R_f$ (proportion of wealth invested is independent of wealth)
\ei
\end{frame}

\bfr\frametitle{Example: Log Utility}
\bi
\im $\tw = w_0[R_f + \pi'(\tR - R_f\iota)]$
\im $\log \tw = \log w_0 + \log [R_f + \pi'(\tR-R_f\iota)]$
\im Optimal $\pi$ maximizes expected value of $\log [R_f + \pi'(\tR-R_f\iota)]$ and is independent of $w_0$.
\ei
\end{frame}


\begin{frame}[plain]
What is the sign of $\mye[u''(\tw)(\tr-R_f)]$?
\bi
\im By the definition of covariance,
$$\mye[u''(\tw)(\tr-R_f)] = \cov(u''(\tw),\tr-R_f) + \mye[u''(\tw)]\mye[\tr-R_f]$$
and the second term is negative.
\im DARA implies the first term is positive and larger in absolute value than the second term.
\ei
\end{frame}

\bfr\frametitle{Proof}
Define $w_f = w_0R_f$ and substitute
\begin{align*}
u''(\widetilde{w}) &= -\alpha (\widetilde{w})u'(\widetilde{w})\\
&= -\alpha (w_f)u'(\widetilde{w}) + \left[\alpha (w_f)- \alpha (\widetilde{w})\right]u'(\widetilde{w})
\end{align*}
in $\mye[u''(\tw)(\tr-R_f)]$ to obtain
$$-\alpha (w_f)\mye\left[(\widetilde{R}-R_f)u'(\widetilde{w})\right] + \mye\left[\left[\alpha (w_f)- \alpha (\widetilde{w})\right](\widetilde{R}-R_f)u'(\widetilde{w})\right]\,.$$
The first term in this expression is zero, due to the FOC.  The second term is positive because
$$\left[\alpha (w_f)- \alpha (\widetilde{w})\right](\widetilde{R}-R_f)$$
is everywhere positive, due to~$\widetilde{w}$ being greater than $w_f$ whenever $\widetilde{R}>R_f$ and the assumption that absolute risk aversion is a decreasing function of wealth.  
\end{frame}


\bfr\frametitle{Exercises}

\end{frame}

\end{document}


\begin{frame}\frametitle{}
\bi
\im
\ei
\end{frame}

\begin{frame}\frametitle{}
\bi
\im
\ei
\end{frame}

\begin{frame}\frametitle{}
\bi
\im
\ei
\end{frame}

\begin{frame}\frametitle{}
\bi
\im
\ei
\end{frame}

\end{document}


\bfr\frametitle{Linear Risk Tolerance}
\bi
\im Linear risk tolerance (LRT): $\tau(w) = A + Bw$.  $B$ is cautiousness parameter.
\im Also called HARA (hyperbolic absolute risk aversion) because absolute risk aversion is $\alpha(w) = 1/(A+Bw)$.
\ei
Examples:
\bi
\im $B=0 \Rightarrow\,$ absolute risk aversion $\,=1/A$.  Called CARA (constant absolute risk aversion).  
\bi
\im Up to a monotone affine transform, all CARA utility functions are $u(w) = - \E^{-\alpha w}$.
\im Here, $\alpha=\,$ absolute risk averison
\ei
\im $A=0
\im CARA $\,\Rightarrow \tau(w) = A = 1/\alpha$.
\im CRRA (constant relative risk aversion) $\,\Rightarrow \tau(w) = Bw = w/\rho$.
\ei
\im Aggregate risk tolerance for individuals $h=1,\ldots,H$: $\sum_{h=1}^H \tau_h(w_h)$
\im Aggregate absolute risk aversion: $1/\sum_{h=1}^H \tau_h(w_h) = 1/ \sum_{h=1}^h (1/\alpha_h(w_h))$
\im $\alpha$ constant nonzero $\,\Rightarrow$ negative exponential utility $u(w) = -\E^{-\alpha w}$.
\im CARA utility $\,\Rightarrow\,$ no wealth effects
\ei
\im CARA
\im CRRA
\bi
\im $u(w) = \log w \Rightarrow \,$ relative risk aversion $\,=1$.
\im $u(w) = \frac{1}{1-\rho}w^{1-\rho} \Rightarrow \,$ relative risk aversion $\,=\rho \neq 1$.
\ei
\im Shifted CRRA
\bi
\im For a constant $\zeta$ and $w>\zeta$, $u(w) = \log (w-\zeta)$.
\im For a constant $\zeta $ and a constant $\rho $ with $\rho  \neq 0$ and $\rho  \neq 1$ and for~$w$ such that $(w-\zeta)/\rho >0$, and including $w=\zeta$ if $\rho<1$,
 $$u(w) = \frac{\rho }{1-\rho }\left(\frac{w-\zeta}{\rho}\right)^{1-\rho }\,.$$
\im When $\rho>0$ (usual case), we can write shifted power as
$$u(w) = \frac{1}{1-\rho }\left(w-\zeta\right)^{1-\rho }\,.$$
\ei 

\ei

\end{frame}

\bfr\frametitle{Quadratic Utility}
Shifted power with $\rho=-1$:
$$u(w) = -\frac{1 }{2}\left(\zeta-w\right)^2$$ 
It is monotone increasing for $w<\zeta$ ($\zeta$ is bliss point).

Mean-Variance Preferences:
$$\mye[u(\tilde{w})] \sim  \zeta \overline{w} - \frac{1}{2}\overline{w}^2 - \frac{1}{2}\var(\widetilde{w})$$

Risk tolerance is $\zeta-w$ which decreases as $w$ increases up to $\zeta$.
\end{frame}

\bfr\frametitle{Higher Moments}
\bi
\im DARA (decreasing absolute risk aversion) $\,\Rightarrow u'''(w)>0$
\im DARA and decreasing absolute prudence $\,\Rightarrow u''''(w)<0$
\im Taylor series expansion:
\begin{multline*}
\mye[u(\tilde{w})] \approx u(\overline{w}) + \overbrace{u'(\ow)\mye[\tw - \ow]}^{\textcolor{red}{0}} + \frac{1}{2}u''(\ow)\mye\lb (\tw-\ow)^2\rb \\+ \frac{1}{6}u'''(\ow)\mye\lb(\tw-\ow)^3\rb + \frac{1}{24} u''''(\ow)\mye\lb(\tw-\ow)^4\rb\,.
\end{multline*}
\im Approximate mean-variance-skewness-kurtosis preferences, liking mean, disliking variance, preferring positive skewness, and disliking kurtosis.
\ei
\end{frame}


\end{document}