\documentclass[xcolor=dvipsnames,10pt]{beamer}
\input{head.tex}
\newcommand{\tu}{\tilde{u}}
\begin{document}
\title{\vskip 0.5in Day 25}
\subtitle{Perpetual Options}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Set-up}
   Single risky asset with price $S$ and constant volatility $\sigma$, single Brownian motion, constant risk-free rate
   
   Dividend paid by risky asset in time period $\D t$ is $qS_t\,\D t$ for constant $q$ (``dividend yield'')
   
   Total return is
   $$\frac{\D S + qS\,\D t}{S} = \frac{\D S}{S} + q\,\D t$$
   Total expected return under RNP is risk-free rate, so
   $$\frac{\D S}{S} = (r-q)\,\D t + \sigma\,\D B^*$$
   for a risk-neutral Brownian motion $B^*$
\end{frame}

\begin{frame}{Perpetual Call}

Perpetual call option with strike $K$

Why exercise?  To capture the dividend.  But the asset price and dividend must be high enough before it is optimal to do so.

An example of a strategy is to pick a number $x$ and exercise the first time $S_t$ gets up to $x$.  The optimal strategy will be of this type.

The problem of finding the optimal exercise time is in the class of problems often called optimal stopping.
\end{frame}

\begin{frame}{Exercise Boundary}
We will first calculate the value if we exercise the first time $S_t$ gets up to $x$ for an arbitrary $x>S_0$.

Let $\tau = \inf\{t \mid S_t \ge x\}$.  This is called the hitting time of $x$.

By the time-homogeneity of $S$, the value at any $t<\tau$ depends only on $S_t$.  Call it $f(S_t)$.  
    
    More formally,
    $$f(s) = \mye^*[\E^{-r\tau}((x-K) \mid S_0=s] = \mye^*[\E^{-r(\tau-t)} (x-K)\mid S_t=s]$$
\end{frame}
\begin{frame}{Fundamental ODE}
    
    The fundamental ODE is 
    $$\frac{\text{drift$^*$ of $f$}}{f} = r$$
    which is
    $$(r-q)Sf' + \frac{1}{2}\sigma^2S^2f'' = rf$$
    Trying a power solution $f(S) = S^\gamma$, we see that $f$ satisfies the ODE if and only if
    $$(r-q)\gamma + \frac{1}{2}\sigma^2\gamma(\gamma-1) = r$$
    The quadratic formula shows that there are two real roots of this equation.  One is negative and the other is greater than 1.  
    

\end{frame}
    
\begin{frame}{General Solution and Boundary Conditions}
Let $\gamma=\,$ absolute value of negative root, and $\beta=\,$ positive root.  The general solution of the ODE is
$$aS^{-\gamma} + bS^\beta$$
for constants $a$ and $b$ that must be determined by boundary conditions.

The value $f$ of the call exercised at the hitting time of $x$ satisfies $f(0)=0$ and $f(x) = x-K$.  The condition $f(0)=0$ implies $a=0$, and the condition $f(x)=x-K$ implies $b=(x-K)x^{-\beta}$.  

The value of the call is
$$f(S_t) = (x-K)\left(\frac{S_t}{x}\right)^\beta$$
    
\end{frame}

\begin{frame}{Optimal Stopping}

To optimize, maximize $(x-K)\left(\frac{S_t}{x}\right)^\beta$ over $x$. 
The factor $S_t^\beta$ is a positive constant and is irrelevant for determining the optimum, so we can maximize
$$(x-K)x^{-\beta} = x^{1-\beta} - Kx^{-\beta}$$

The FOC is
$$(1-\beta)x^{-\beta} + \beta Kx^{-\beta-1}=0$$
Equivalently,
$$(1-\beta)x + \beta K=0$$
So, 
$$x^* = \frac{\beta}{\beta-1}K$$
    
\end{frame}

\begin{frame}{Perpetual Put}

Recall that the general solution of the ODe is $f(s) = as^{-\gamma} + bs^\beta$.

For a put, we exercise the first time $S_t$ drops to a boundary $x$.  
The boundary conditions for a put are $f(\infty)=0$, and $f(x)=K-x$.  The condition $f(\infty)=0$ implies $b=0$.  The condition $f(x)=K-x$ implies $a=(K-x)x^{\gamma}$.

So, the put value is 
$$f(S_t) = (K-x)\left(\frac{x}{S_t}\right)^\gamma$$
The FOC for maximizing over $x$ is
$$\gamma K x^{\gamma-1} - (1+\gamma)x^\gamma =0$$
Maximizing over $x$ yields $x^*=\gamma K/(1+\gamma)$.

    
\end{frame}
\end{document}