\documentclass[10pt]{beamer}

\usetheme[progressbar=foot]{metropolis}
\usepackage{appendixnumberbeamer}

\usepackage{booktabs}
\usepackage[scale=2]{ccicons}

\usepackage{pgfplots}
\usepgfplotslibrary{dateplot}

\usepackage{xspace}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=, urlcolor=blue}

\DeclareMathOperator{\stdev}{stdev}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\corr}{corr}
\DeclareMathOperator{\prob}{prob}
\DeclareMathOperator{\n}{n}
\DeclareMathOperator{\N}{N}
\DeclareMathOperator{\Cov}{Cov}

\newcommand{\hlf}{\frac{1}{2}}
\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\im}{\item}
\newcommand{\D}{\mathrm{d}}
\newcommand{\E}{\mathrm{e}}
\newcommand{\mye}{\ensuremath{\mathsf{E}}}
\newcommand{\myreal}{\ensuremath{\mathbb{R}}}
\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\bfr}{\begin{frame}}
\newcommand{\efr}{\end{frame}}
\newcommand{\bas}{\begin{align*}}
\newcommand{\eas}{\end{align*}}

\setbeamertemplate{frame footer}{PCA and Factor Models, BUSI 521/ECON 505}

\title{Principal Components and Factor Models}

\date{}
\author{Kerry Back\\
BUSI 521/ECON 505\\
Rice University\\[1em]
\small\href{https://colab.research.google.com/github/kerryback/busi521/blob/main/PCA_Factors.ipynb}{Open companion notebook in Colab}}


\begin{document}

\maketitle

\section{Factor Models}

\begin{frame}{Setup}
Let $R_t$ be an $N$--vector of excess returns with mean $\mu$ and covariance matrix $\Sigma$.

A $K$--factor model is
$$R_t = \mu + Bf_t + e_t$$
\bi
\im $f_t$ is a $K$--vector of factors (zero mean, $\Cov(f_t)=I$)
\im $B = \Cov(R_t, f_t)$ is the $N \times K$ loading matrix
\im $e_t$ is uncorrelated with $f_t$
\ei
\end{frame}

\begin{frame}{What Makes It a Factor Model?}
The key restriction: the elements of $e_t$ are \alert{mutually uncorrelated}.

So $\Cov(e_t) = D$ for a diagonal matrix $D$.

This implies
$$\Sigma = BB' + D$$

All return covariances are explained by common factor exposures.  Residual risks are idiosyncratic.
\end{frame}

\section{Factor Analysis}

\begin{frame}{Rotational Indeterminacy}
The decomposition $\Sigma = BB' + D$ does not uniquely determine $B$.

For any $K \times K$ orthogonal matrix $Q$ (so $QQ'=I$), define $\tilde{B} = BQ$.  Then
$$\tilde{B}\tilde{B}' = BQQ'B' = BB'$$

So $\tilde{B}$ and $\tilde{f}_t = Q'f_t$ give the same covariance structure.  The loading matrix is only identified up to an orthogonal rotation.
\end{frame}

\begin{frame}{The Identification Restriction}
To pin down $B$, maximum likelihood factor analysis imposes:
$$B'D^{-1}B \quad \text{is diagonal}$$

\alert{Interpretation.}  In the pre-whitened space, define $\tilde{B}=D^{-1/2}B$.  Then
$$\tilde{B}'\tilde{B} = B'D^{-1}B$$
So the restriction says the columns of $\tilde{B}$ (the whitened loadings) are orthogonal.  This is natural because $B'D^{-1}B$ appears in the GLS formula for factor inference.
\end{frame}

\begin{frame}{Why It Resolves Indeterminacy}
A $K\times K$ orthogonal matrix $Q$ has $K(K{-}1)/2$ free parameters (one angle per pair of factors).  The restriction ``$B'D^{-1}B$ is diagonal'' kills the $K(K{-}1)/2$ off-diagonal elements---a perfect match.

\bi
\im \alert{Existence.}  Given any $B$, the symmetric matrix $B'D^{-1}B$ has an eigendecomposition $Q\Lambda Q'$.  Setting $\tilde{B}=BQ$ gives $\tilde{B}'D^{-1}\tilde{B} = \Lambda$ (diagonal).
\im \alert{Uniqueness.}  If the diagonal entries of $\Lambda$ are distinct, the only orthogonal $Q$ satisfying $Q'\Lambda Q = \Lambda$ is a signed permutation matrix (sign flips and reordering of factors), which is a trivial relabeling.
\ei

\medskip
We also need enough data moments.  The model has $NK + N - K(K{-}1)/2$ free parameters in $B$ and $D$, which must not exceed the $N(N{+}1)/2$ distinct elements of $\Sigma$.
\end{frame}

\begin{frame}{ML Estimation}
Given the identification restriction, ML estimation recovers $\hat{B}$ and $\hat{D}$ from the sample covariance matrix.

But we still need the factor realizations $f_t$ at each date.
\end{frame}

\begin{frame}{Inferring the Factors}
At each date $t$, the model says
$$\underbrace{R_t - \mu}_{N\times 1} = \underbrace{B}_{N\times K}\;\underbrace{f_t}_{K\times 1} + \underbrace{e_t}_{N\times 1}$$

This is a \alert{cross-sectional regression}: $N$ observations (one per asset), $K$ unknowns (the factor values $f_t$), and $B$ plays the role of the design matrix.

\bi
\im OLS estimate: $\;\hat{f}_t = (B'B)^{-1}B'(R_t - \mu)$
\im GLS estimate: $\;\hat{f}_t = (B'D^{-1}B)^{-1}B'D^{-1}(R_t - \mu)$
\ei

GLS is preferred because $\Cov(e_t)=D$ is heteroskedastic (different assets have different idiosyncratic variances).
\end{frame}

\begin{frame}{Why GLS Produces Uncorrelated Residuals}
The factor model's key restriction: $\Cov(e_t)=D$ is diagonal---all correlations come from common factors.

\alert{Pre-whitening.}  Define $\tilde{R}_t = D^{-1/2}(R_t - \mu)$ and $\tilde{B} = D^{-1/2}B$.  Then
$$\tilde{R}_t = \tilde{B}f_t + \tilde{e}_t\,,\qquad \Cov(\tilde{e}_t)=I$$

OLS on the whitened model gives exactly the GLS formula:
$$\hat{f}_t = (\tilde{B}'\tilde{B})^{-1}\tilde{B}'\tilde{R}_t = (B'D^{-1}B)^{-1}B'D^{-1}(R_t - \mu)$$

Because $\tilde{e}_t$ has identity covariance (iid errors), OLS correctly separates factors from uncorrelated noise in the whitened space.

\medskip
Skipping the whitening (i.e., using OLS on the raw data) ignores heteroskedasticity, letting idiosyncratic noise leak across assets and creating spurious residual correlations.
\end{frame}

\begin{frame}{GLS Creates Traded Factors}
The GLS formula
$$f_t = (\hat{B}'\hat{D}^{-1}\hat{B})^{-1}\hat{B}'\hat{D}^{-1}(R_t - \hat{\mu})$$
shows that each factor is a \alert{linear combination of returns}---a portfolio return (up to a constant).

\bi
\im The companion notebook confirms this: regressing any inferred factor on the $N$ returns gives $R^2=1$ (exact spanning).
\im This exact spanning is what makes the FMP construction work.  The regression coefficients, normalized to sum to one, are the portfolio weights.
\ei

\medskip
In python: \texttt{sklearn.decomposition.FactorAnalysis}

\smallskip
{\small \href{https://colab.research.google.com/github/kerryback/busi521/blob/main/PCA_Factors.ipynb}{$\rightarrow$ Open companion notebook in Colab}}
\end{frame}

\section{Principal Components}

\begin{frame}{Eigenvectors of the Covariance Matrix}
Let $C$ be the matrix of eigenvectors of $\Sigma$, ordered by eigenvalue (largest first).

Partition $C = (C_1 \mid C_2)$ where $C_1$ contains the first $K$ eigenvectors.

The principal component factors are
$$f_t = (R_t - \hat{\mu})\,C_1$$

The eigenvectors are orthonormal: $C'C = CC' = I$.
\end{frame}

\begin{frame}{PCA as a Factor Model}
Since $(f_t,\, \text{extras}_t) = (R_t - \hat{\mu})\,C$ and $CC'=I$:
$$R_t = \hat{\mu} + f_t\, C_1' + \text{extras}_t\, C_2'$$
\bi
\im The loading matrix is $C_1'$ (the transposed eigenvectors)
\im The ``residuals'' are $\text{extras}_t \cdot C_2'$
\im Factors and residuals are orthogonal
\ei

PCA does not force residuals to be uncorrelated.  Instead, it makes residuals \alert{small} by dropping eigenvectors with small eigenvalues.
\end{frame}

\begin{frame}{How Much Variance Is Explained?}
Each eigenvalue $\lambda_i$ equals the variance of the $i$th principal component.

The fraction of total variance explained by $K$ components is
$$\frac{\lambda_1 + \cdots + \lambda_K}{\lambda_1 + \cdots + \lambda_N}$$

For the 25 size/BM portfolios, the first eigenvalue alone accounts for about 82\% of total variance.  Three components capture about 93\%.

In python: \texttt{sklearn.decomposition.PCA}
\end{frame}

\section{Factor-Mimicking Portfolios}

\begin{frame}{Why Factor-Mimicking Portfolios?}
The raw factors (from FA or PCA) are linear combinations of returns, but the weights don't sum to one.

A factor-mimicking portfolio (FMP) rescales the weights so $\iota'w = 1$.

The FMP excess return is then a \alert{traded} excess return that tracks the factor.
\end{frame}

\begin{frame}{Constructing FMPs: PCA}
For PCA, the weights are the eigenvector entries.  To make an FMP:
$$w_i = \frac{C_{1,i}}{\iota' C_{1,i}}$$

That is, divide each eigenvector's entries by their sum.

This gives $K$ portfolios whose excess returns are the PCA factor-mimicking portfolios.
\end{frame}

\begin{frame}{Constructing FMPs: Factor Analysis}
For factor analysis, regress each factor $f_t^{(i)}$ on a constant and the $N$ excess returns:
$$f_t^{(i)} = \alpha + w'R_t + \text{error}$$

The regression coefficients $w$ give the spanning weights.  Normalize so $\iota'w=1$.

The FMP return is $w'R_t$.

The $R^2$ is 1 (exact spanning) because the factors were constructed from the returns via the GLS formula.
\end{frame}

\section{Optimization Perspectives}

\begin{frame}{PCA $=$ Maximum-Variance Portfolios}
The first principal component solves
$$\max_{w}\; \hlf w'\Sigma w \quad \text{subject to}\quad w'w = 1$$
The FOC is $\Sigma w = \lambda w$\,---\,the eigenvector equation.

The second component solves the same problem with the additional constraint $w'w_1 = 0$.

Successive components are maximum-variance portfolios orthogonal to all previous ones.
\end{frame}

\begin{frame}{FA $=$ Minimum Residual-Variance Portfolios}
The factor-analysis FMPs solve
$$\min_{w}\; \hlf w'Dw \quad \text{subject to}\quad B'w = e_i$$

This minimizes idiosyncratic risk while targeting a unit loading on factor $i$ and zero loadings on all other factors.

The FOC gives $w = D^{-1}B(B'D^{-1}B)^{-1}e_i$, and the resulting portfolio returns reproduce exactly the GLS factor estimates.
\end{frame}

\section{Sharpe Ratios}

\begin{frame}{Maximum Sharpe Ratios}
Given $K$ excess-return factors with mean $\mu_f$ and covariance $\Sigma_f$, the maximum Sharpe ratio from combining them is
$$\text{SR}^* = \sqrt{\mu_f'\Sigma_f^{-1}\mu_f}$$

This is the Sharpe ratio of the tangency portfolio of the $K$ factors.

Using monthly excess returns on the 25 Fama--French size/BM sorted portfolios from 1980--2021, with $K=3$:
\bi
\im Fama--French factor SR $\approx 0.178$
\im FA mimicking-portfolio SR $\approx 0.183$
\im PCA mimicking-portfolio SR $\approx 0.177$
\ei

All three are similar---the statistical factors capture roughly the same pricing information as the Fama--French factors.
\end{frame}

\begin{frame}{Summary}
\bi
\im \alert{Factor analysis}: ML estimation of loadings and diagonal residual covariance; factors inferred by GLS.  FMPs minimize idiosyncratic risk.
\im \alert{PCA}: Eigenvectors of the covariance matrix; factors are maximum-variance linear combinations.  FMPs are rescaled eigenvectors.
\im Both methods produce traded factor-mimicking portfolios by normalizing weights to sum to one.
\im For the 25 size/BM portfolios, three statistical factors deliver Sharpe ratios comparable to Fama--French.
\ei
\end{frame}

\end{document}
